OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-01-26 paper/paper.Rmd
Introduce algorithms

- **TODO: broad context; importance of amplicon sequencing**
- **TODO: describe reference vs de novo clustering. trade-offs. machine learning use-case.**
- **TODO: mention OptiClust and vsearch.**

@@

To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by _de novo_ clustering the Greengenes
database (v13_8_99), Silva non-redundant database (v132), and the Ribosomal
Database Project (RDP; v16). In open-reference mode, OTU quality was similar
between fitting the datasets to reference OTUs with OptiFit and clustering the
datasets _de novo_ with OptiClust. This held true for all datasets and reference
databases. However, in closed-reference mode, OTU quality was slightly worse
when fitting to greengenes and silva, and much worse when fitting to RDP as
compared to OptiClust. No more than half of query sequences were fit to
reference OTUs in closed-reference mode across any dataset/database combination.
OptiFit was able to fit more query sequences to reference OTUs created with the
greengenes and silva databases than with RDP. In terms of runtime,
closed-reference OptiFit outperformed OptiClust, while OptiClust outperformed
open-reference OptiFit.

@@

@@

When fitting to public databases, OTU quality dropped in closed-reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP. We recommend users who require
an independent reference database opt for Greengenes or Silva instead. Since
OptiClust performs faster than open-reference OptiFit and creates higher quality
OTUs than closed-reference OptiFit with the database strategy, we recommend
using OptiClust rather than fitting to a database where possible.

>>>>

Amplicon sequencing has become a mainstay of microbial ecology and
host-associated microbiome research. Researchers can affordably generate
millions of sequences to characterize the composition of hundreds of samples
from culture-independent microbial communities. In a typical analysis pipeline,
16S rRNA gene sequences are assigned to Operational Taxonomic Units (OTUs) to
facilitate comparison of taxonomic composition between communities. A distance
threshold of 3% (or sequence similarity of 97%) is commonly used to cluster
sequences into OTUs based on either a reference database or pairwise comparisons
of the sequences within the dataset. The method chosen for clustering affects
the quality of OTU assignments and thus may impact downstream analyses of 
community composition [@westcott_opticlust_2017; @schloss_application_2016; @westcott_novo_2015].

Closed-reference methods assign sequences to a set of pre-made OTUs generated
from reference sequences. If a query sequence is not within the distance
threshold to any of the reference sequences, it is discarded. _De novo_ methods
cluster sequences based on their distance to each other, without the use of an
external reference. Open-reference methods first perform closed-reference
clustering, then any sequences that cannot be assigned to reference OTUs are
clustered _de novo_ to create additional OTUs. While reference-based clustering
is generally fast, it is limited by the diversity represented in the reference
database. _De novo_ clustering overcomes the limitations of reference databases
by considering only sequences in the dataset, but is more computationally
intensive and generates different OTU assignments when new sequences are
introduced.

- **TODO: clustering methods. describe reference vs de novo clustering. trade-offs. machine learning use-case.**
- **TODO: mention OptiClust and vsearch.**
- **TODO:** gap in knowledge/ problem statement. acknowledge knock against OTUs, that they're not reproducible across studies. idea of ref-clustering is you can use existing OTUs to compare data across studies. can we make a ref-based method that's as good as _de novo_.
- **TODO:** in this study, we fix that problem.

@@

- TODO: separate paragraphs for closed & open reference clustering. weave in vsearch within those paragraphs. mention median mcc values.
- TODO: avoid slightly/much -- use numbers, e.g. X% better/worse.
- TODO: put the comparison at the beginning of the sentence, so you don't have to know what it is at the end.

To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by _de novo_ clustering the Greengenes
database (v13_8_99), Silva non-redundant database (v132), and the Ribosomal
Database Project (RDP; v16). In open-reference mode, OTU quality was similar
between fitting the datasets to reference OTUs with OptiFit and clustering the
datasets _de novo_ with OptiClust. This held true for all datasets and reference
databases. However, in closed-reference mode, OTU quality was slightly worse
when fitting to greengenes and silva, and much worse when fitting to RDP as
compared to OptiClust. No more than half of query sequences were fit to
reference OTUs in closed-reference mode across any dataset/database combination.
**TODO: specify exact numbers for fraction mapped.**
OptiFit was able to fit more query sequences to reference OTUs created with the
greengenes and silva databases than with RDP. In terms of runtime,
closed-reference OptiFit outperformed OptiClust, while OptiClust outperformed
open-reference OptiFit.

@@

- **TODO:** for these data, we don't see a compelling reason to use reference-based clustering over _de novo_. it was supposed to speed things up. the reason to do reference-based is if you like the ref OTUs -- e.g. for ML or downstream tools e.g. PiCrust?
- **TODO:** highlight difference between what we're doing and what previously was done.
  others use single ref seq to define an otu, while we use all the ref & query seqs to define an otu. highlight why ours is better than previous methods.

@@

When fitting to public databases, OTU quality dropped in closed-reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP. We recommend users who require
an independent reference database opt for Greengenes or Silva instead. Since
OptiClust performs faster than open-reference OptiFit and creates higher quality
OTUs than closed-reference OptiFit with the database strategy, we recommend
using OptiClust rather than fitting to a database where possible.
(**Pat: if you don't have breadth, closed-ref will suck.**)
